\section{Rebalancing}
\label{sec:rebalancing}

Rebalancing moves the durable metadata for a shard range from one \ipto{}
instance to another after a placement change. \vannak{} provides the building
blocks for this, while the surrounding automation remains future work.

\subsection{Motivation}

When the placement map changes, a data individual's new writes go to its new
owner, but its older durable metadata still lives on the previous owner. Two
mechanisms cover this gap: a read-time fallback that is available immediately,
and an explicit replay that physically moves the data.

\subsection{Query Fallback Approach}

The first line of defense is query fallback. Because
\texttt{ClusterControlState} keeps a bounded history of placement maps,
\texttt{resolve\_with\_fallback} returns the current owner first and then prior
owners. A reader therefore finds metadata written before a reconfiguration
without any physical movement, as long as the change is within the five-epoch
history window (Section~\ref{sec:topology}).

\begin{vnote}
Query fallback is sufficient for short-lived reconfigurations. It does not move
data, so it cannot replace explicit rebalancing for long-lived data or for
permanently retiring an \ipto{} instance.
\end{vnote}

\subsection{Shard-Aware Payload}

Physical rebalancing is possible because every \texttt{IptoWritePayload} records
its \texttt{shard\_id}. The encoded payload in an outbox segment therefore
carries enough information to select exactly the records belonging to a shard
range, without consulting any external index.

\subsection{Shard-Range Replay}

\texttt{replay\_metadata\_outbox\_segment\_for\_shard\_range} scans a segment and
returns an outbox containing only the pending entries whose \texttt{shard\_id}
falls within an inclusive range.

\begin{lstlisting}
pub fn replay_metadata_outbox_segment_for_shard_range(
    path: impl AsRef<Path>,
    start: DataIndividualShardId,
    end: DataIndividualShardId,
) -> Result<MetadataOutbox, MetadataOutboxStorageError> {
    let mut reader = SegmentReader::open(path)?;
    let mut outbox = MetadataOutbox::new();
    while let Some(record) = reader.read_next()? {
        let payload = IptoWritePayload::decode(&record.payload)?;
        if payload.shard_id >= start && payload.shard_id <= end {
            outbox.enqueue_with_offset(payload, Some(record.offset));
        }
    }
    Ok(outbox)
}
\end{lstlisting}

\subsection{rebalance\_shard\_range\_to()}

\texttt{rebalance\_shard\_range\_to} ties replay and delivery together. It
replays the shard range from a segment, then drains the resulting pending
entries through a writer connected to the new \ipto{} target. Because delivery
goes through the idempotent writer, partially completed rebalances are safe to
re-run.

\begin{lstlisting}
pub fn rebalance_shard_range_to(
    segment_path: impl AsRef<Path>,
    start: DataIndividualShardId,
    end: DataIndividualShardId,
    writer: &mut impl IptoWriter,
    max_attempts: usize,
) -> Result<MetadataOutboxRebalanceSummary, MetadataOutboxStorageError>
\end{lstlisting}

The returned \texttt{MetadataOutboxRebalanceSummary} reports
\texttt{entries\_found}, \texttt{attempted}, \texttt{acknowledged}, and
\texttt{failed}. The integration test \texttt{rebalancing\_delivers\_shard\_range\_to\_new\_target}
writes two payloads with different shard ids and confirms that rebalancing one
shard range delivers only the matching entry.

\subsection{When to Rebalance}

Use explicit rebalancing rather than query fallback when:

\begin{itemize}
\item an \ipto{} instance is being permanently retired;
\item the placement change is older than the five-epoch fallback window;
\item read latency from fanning out across prior owners is unacceptable;
\item compliance requires data to physically reside on its new owner.
\end{itemize}

\subsection{Not Yet Implemented}

The primitives above exist and are tested, but the surrounding orchestration is
not yet built. There is currently no automatic trigger that detects a placement
change and schedules rebalancing, no progress checkpoint specific to a
rebalance, and no coordinated cut-over that fences the old target. Operators
invoke \texttt{rebalance\_shard\_range\_to} explicitly; placement-driven,
checkpointed, cluster-coordinated rebalancing is planned future work that builds
on writer leases and outbox checkpoints.

\endinput
