\section{Appendix}
\label{sec:appendix}

This appendix collects reference material: the command-line interface,
environment variables, infrastructure files, the wire and storage formats, the
repository layout, and a glossary.

\subsection{Functionality Status Matrix}

This matrix cross-references the operating model described in this manual with
the functionality currently present in the repository.

\begin{description}
\item[Implemented] compiled and exercised by unit tests or the normal build.
\item[Implemented primitive] the core type or helper exists, but no daemon or
  orchestration loop wires it into an end-to-end service path.
\item[Implemented service boundary] wired through the in-process
  \texttt{VannakService}; still not an external daemon or \sitas{} runtime.
\item[Feature-gated] compiled only with the named Cargo feature.
\item[Test/runbook] available through tests, examples, scripts, or operator
  commands rather than as a continuously running service.
\item[Planned] documented design direction; not implemented in this repository
  yet.
\end{description}

\begin{longtable}{p{0.24\textwidth}p{0.20\textwidth}p{0.48\textwidth}}
\hline
Capability & Status & Code or manual reference \\
\hline
\endfirsthead
\hline
Capability & Status & Code or manual reference \\
\hline
\endhead
\hline
\endfoot
Durga-compatible process-event mirror & Implemented &
\texttt{durga.rs}, \texttt{ingest.rs}, and \texttt{HotIndex::ingest}; tested
through the Durga conversion test. \\

Live Durga Kafka ingestion & Feature-gated &
\texttt{kafka\_ingest.rs} defines decoded Kafka process-event records with
topic, partition, and offset identity, journals them when configured, feeds the
\sitas{} mailbox path, retains pending records across backpressure retries, and
returns the next committable offset.
\texttt{kafka\_client.rs} and \texttt{vannak-kafka-ingest} are compiled with
\texttt{kafka-client}; they use \texttt{rdkafka}, subscribe to process-event
topics, decode either Vannak binary process-event payloads or Durga JSON monitor
events, feed Sitas workers, pause Kafka partitions when Sitas mailboxes are
full, resume them after admission, record assignment/revocation callbacks in
owned snapshots, and commit offsets after admission. If a partition is revoked
while it still has an uncommitted pending record, the consumer returns an
explicit error instead of committing the offset. Live Durga schema verification,
richer operational supervision, and live Durga compatibility coverage remain
planned.
\\

Kafka broker smoke test & Test/runbook &
\texttt{docker-compose.kafka.yml} starts a single Redpanda broker.
\texttt{tests/kafka\_integration.rs} is enabled with
\texttt{VANNAK\_KAFKA\_INTEGRATION=1}; it produces a Durga JSON process event,
consumes it through \texttt{KafkaProcessConsumer}, and verifies that \sitas{}
workers reduce the event into the hot index. \\

Process-state reduction & Implemented &
\texttt{ProcessInstanceState::apply} and \texttt{HotIndex}; maintains current
status, activity states, durations, metadata references, and owned snapshots. \\

Single-node hot index & Implemented &
\texttt{HotIndex} indexes events by id, process instance, pipeline, and
metadata reference. It is intentionally non-concurrent shard-local state. \\

Single-node service orchestration & Implemented service boundary &
\texttt{VannakService} wires \texttt{HotIndex}, \texttt{IptoPlacementMap},
\texttt{IptoMapping}, and \texttt{SegmentBackedMetadataOutbox} for process
ingest, durable metadata capture, lease-gated target drains, checkpoint lookup,
and owned snapshots. \\

Shard-local runtime model & Implemented &
\texttt{ShardLocalRuntime} owns one \texttt{HotIndex} per
\texttt{LogicalShardId}, routes process-instance-owned events and queries to
one shard, fans out pipeline and impact queries, and returns owned aggregate
snapshots. \\

Bounded local ingest queues & Implemented &
\texttt{BoundedIngestRuntime} adds bounded per-shard process-event queues,
explicit \texttt{RuntimeError::QueueFull} backpressure, shard drain summaries,
and owned queue-depth snapshots. \\

\sitas{} executor integration & Feature-gated &
\texttt{SitasShardRuntime} is compiled with \texttt{sitas-runtime}. It starts a
\sitas{} \texttt{ShardedExecutor}, stores one \texttt{HotIndex} in
\texttt{ShardLocal} state per shard, routes direct process-event ingest through
executor submissions, and exposes owned executor/hot-index snapshots. \\

\sitas{} process-event mailboxes & Feature-gated &
\texttt{SitasShardRuntime} creates a bounded
\texttt{ShardMailboxSet<PipelineEvent>}; \texttt{try\_submit} returns explicit
backpressure on full mailboxes, \texttt{submit} waits for capacity, manual
\texttt{drain\_shard}/\texttt{drain\_all} remains available, and
\texttt{start\_mailbox\_workers} starts one continuous receiver per Sitas shard
to feed the owning shard-local \texttt{HotIndex}. \\

Typed query primitives & Implemented &
\texttt{query.rs} plus \texttt{HotIndex::process\_instance},
\texttt{pipeline\_instances}, \texttt{events}, \texttt{impact},
\texttt{process\_instances\_by\_status}, and
\texttt{events\_in\_time\_range}. Data-individual metadata queries are exposed
through \texttt{DataIndividualMetadataQuery}, \texttt{ProcessMetadataQuery}, and
\texttt{ActivityMetadataQuery}. \\

Local query fanout & Implemented &
\texttt{ShardLocalRuntime::pipeline\_instances} and
\texttt{ShardLocalRuntime::impact}, plus status and time-range query methods,
scatter across local shards and merge owned results with global limits.
Cross-node cluster query fanout is still planned. \\

Data-individual provenance model & Implemented &
\texttt{DataIndividualMetadataEvent}, \texttt{PassiveMetadata},
\texttt{ActiveMetadata}, \texttt{MetadataOperation}, and deterministic
\texttt{IdempotencyKey} derivation. \\

Local provenance lookup index & Implemented &
\texttt{DataProvenanceIndex} idempotently indexes captured metadata events by
data individual, process instance, and optional activity. \texttt{VannakService}
updates it after durable metadata enqueue and exposes owned local query results.
\\

Domain metadata placement key & Implemented &
\texttt{DataIndividualShardId::from\_data\_individual}; explicitly separate
from \sitas{} executor placement. \\

Ipto placement ring and overrides & Implemented &
\texttt{IptoPlacementRing}, \texttt{IptoPlacementRange}, and
\texttt{IptoPlacementMap}; epoch validation and five-epoch fallback history
live in \texttt{ClusterControlState}. \\

Ipto field mapping and payload codec & Implemented &
\texttt{IptoMapping}, \texttt{IptoWritePayload::from\_event},
\texttt{encode}, and \texttt{decode}. Only mapped fields are written. \\

In-memory metadata outbox & Implemented &
\texttt{MetadataOutbox} supports duplicate detection, pending delivery,
acknowledgement, failure state, retry, and snapshots. \\

Segment-backed metadata outbox & Implemented service boundary &
\texttt{SegmentBackedMetadataOutbox::enqueue\_durable} appends and syncs before
exposing a payload as pending. \texttt{VannakService::capture\_metadata\_event}
wires placement, mapping, and durable enqueue into one local capture path. \\

Append-only segment format & Implemented &
\texttt{SegmentWriter}, \texttt{SegmentReader}, \texttt{SegmentManifest}, magic
validation, record checksums, and seal/manifest support. \\

Raw process-event segment history & Implemented service boundary &
\texttt{ProcessEventJournal::append\_durable} validates, appends, and syncs
encoded \texttt{PipelineEvent} records. \texttt{replay\_process\_event\_segment}
returns owned \texttt{JournaledPipelineEvent} values, and
\texttt{VannakService::ingest\_process\_event\_durable} journals before local
hot-index reduction. \texttt{vannak-kafka-ingest} can use the same journal
before committing Kafka offsets. \\

Outbox replay after checkpoint & Implemented service boundary &
\texttt{replay\_metadata\_outbox\_segment\_after} rebuilds pending state after
a checkpoint offset and returns \texttt{MetadataOutboxReplaySummary};
\texttt{VannakService::replay\_metadata\_outbox\_after} exposes the same local
recovery helper. \\

Shard-range rebalance replay & Implemented primitive &
\texttt{replay\_metadata\_outbox\_segment\_for\_shard\_range} and
\texttt{rebalance\_shard\_range\_to} exist and are tested. Automatic
placement-change orchestration is planned. \\

Ipto writer adapter & Feature-gated &
\texttt{IptoRepoWriter} is compiled with \texttt{ipto-writer}; it configures
the built-in PROV-O SDL and writes units through \texttt{ipto\_rust}. \\

PostgreSQL-backed Ipto integration test & Test/runbook &
\texttt{tests/vannak\_integration.rs} runs only when
\texttt{VANNAK\_PG\_INTEGRATION=1}; otherwise it returns early. \\

Raft-replicable control state & Implemented &
\texttt{ClusterControlState} applies node, placement, lease, sealed-segment,
outbox-checkpoint, and checkpoint-manifest commands. \\

Raft state-machine adapter & Feature-gated &
\texttt{ClusterStateMachine} is compiled with \texttt{raft}; commands,
snapshots, restore, and seven read queries are JSON encoded. \\

\texttt{vannak-node} control-plane binary & Feature-gated &
Compiled with \texttt{node}; hosts a \graft{} node over TCP with DNS SRV
discovery and file-backed \raft{} storage. It does not ingest data events. \\

Target-aware writer lease enforcement & Implemented service boundary &
\texttt{WriterLease} is stored and epoch-validated in control state.
\texttt{VannakService::drain\_metadata\_for\_target\_if\_lease\_held} checks the
lease holder before draining only that target. Background lease renewal and
writer task scheduling remain future work. \\

Local metadata outbox crash recovery & Implemented service boundary &
\texttt{SegmentBackedMetadataOutbox::recover\_after} replays records after a
checkpoint and reopens the validated segment for append.
\texttt{VannakService::recover\_after} rebuilds the single-node service around
that recovered outbox. Automatic discovery of owned segments and background
delivery resumption at daemon startup remain future work. \\

Owned observability snapshots & Implemented primitive &
\texttt{HotIndexSnapshot}, \texttt{MetadataOutboxSnapshot},
\texttt{SegmentBackedMetadataOutboxSnapshot}, replay summaries, and rebalance
summaries are owned values. No metrics or HTTP health endpoint exists. \\

Load generator & Test/runbook &
\texttt{vannak-load} simulates process and metadata events in-process; with
\texttt{ipto-writer} it can write through \ipto{}, but it is not the production
service loop. \\
\end{longtable}

\subsection{CLI Reference}

The \texttt{vannak-node} binary (built with the \texttt{node} feature) accepts
the following forms:

\begin{table}[htbp]
\centering
\begin{tabular}{ll}
\hline
Command & Purpose \\
\hline
\shellcmd{vannak-node <host> <port> <peer-id> <data-dir> [peers...]} & start with an explicit peer list \\
\shellcmd{vannak-node --srv <service> <host> <port> <peer-id> <data-dir>} & start with DNS SRV discovery \\
\shellcmd{vannak-node probe <host> <port>} & probe a node for a cluster summary \\
\hline
\end{tabular}
\caption{\texttt{vannak-node} command forms.}
\end{table}

A peer specification has the form \texttt{peer-id@host:port}. The default port
when parsing fails is 10081.

\subsection{Environment Variables}

\begin{table}[htbp]
\centering
\begin{tabular}{ll}
\hline
Variable & Meaning \\
\hline
\texttt{IPTO\_PG\_HOST} & \ipto{} PostgreSQL host (default \texttt{localhost}) \\
\texttt{IPTO\_PG\_PORT} & \ipto{} PostgreSQL port (default \texttt{5432}) \\
\texttt{IPTO\_PG\_USER} & \ipto{} PostgreSQL user (default \texttt{repo}) \\
\texttt{IPTO\_PG\_PASSWORD} & \ipto{} PostgreSQL password (default \texttt{repo}) \\
\texttt{IPTO\_PG\_DATABASE} & \ipto{} PostgreSQL database (default \texttt{repo}) \\
\texttt{VANNAK\_PG\_INTEGRATION} & set to \texttt{1} to enable the integration test \\
\texttt{VANNAK\_PG\_PORT} & host port mapping for the test PostgreSQL \\
\texttt{VANNAK\_KAFKA\_INTEGRATION} & set to \texttt{1} to enable the Kafka smoke test \\
\texttt{VANNAK\_KAFKA\_BROKERS} & Kafka broker list for the smoke test (default \texttt{localhost:19092}) \\
\texttt{VANNAK\_KAFKA\_PORT} & host port mapping for the Redpanda Kafka API \\
\texttt{POSTGRES\_USER} & test PostgreSQL role (\texttt{repo}) \\
\texttt{POSTGRES\_PASSWORD} & test PostgreSQL password (\texttt{repo}) \\
\texttt{POSTGRES\_DB} & test PostgreSQL database (\texttt{repo}) \\
\hline
\end{tabular}
\caption{All recognized environment variables.}
\end{table}

\subsection{Docker Compose Reference}

\begin{description}
\item[\filepath{docker-compose.test.yml}] PostgreSQL 16 with the \ipto{} schema,
  procedures, and boot data mounted into \filepath{/docker-entrypoint-initdb.d};
  used by the integration test.
\item[\filepath{docker-compose.kafka.yml}] Redpanda broker exposing a
  Kafka-compatible API on \texttt{localhost:19092}; used by
  \texttt{tests/kafka\_integration.rs}.
\item[\filepath{docker-compose.cluster.yml}] CoreDNS plus three
  \texttt{vannak-node} containers on the \texttt{10.77.0.0/24} network, each
  listening on port 10081 with DNS SRV discovery.
\end{description}

\subsection{DNS Zone File}

The zone served by CoreDNS from \filepath{dns/vannak.local.db}:

\begin{lstlisting}[numbers=none]
$ORIGIN vannak.local.
$TTL 5
@       IN SOA  ns.vannak.local. hostmaster.vannak.local. (2026060601 5 5 60 5)
        IN NS   ns.vannak.local.

ns      IN A    10.77.0.10

vannak-1 IN A    10.77.0.11
vannak-2 IN A    10.77.0.12
vannak-3 IN A    10.77.0.13

_vannak._tcp IN SRV 10 10 10081 vannak-1.vannak.local.
_vannak._tcp IN SRV 10 10 10081 vannak-2.vannak.local.
_vannak._tcp IN SRV 10 10 10081 vannak-3.vannak.local.
\end{lstlisting}

\subsection{Port Assignments}

\begin{table}[htbp]
\centering
\begin{tabular}{ll}
\hline
Port & Use \\
\hline
\texttt{10081} & \raft{} transport and \texttt{probe} (all nodes) \\
\texttt{53} & CoreDNS (container-internal) \\
\texttt{5432} & PostgreSQL (test backend) \\
\hline
\end{tabular}
\caption{Port assignments.}
\end{table}

\subsection{Ipto SQL Schema}

The integration backend loads three external SQL files from
\filepath{../ipto/shared/db/postgresql}: \filepath{schema.sql} (tables),
\filepath{procedures.sql} (stored procedures), and \filepath{boot.sql} (boot
data). \vannak{} treats these as the authoritative \ipto{} schema and does not
redefine them. The relevant facts for \vannak{} are that boot data seeds tenant
id 1 (\texttt{SCRATCH}), and that provenance is stored as units carrying a
\texttt{tenantid}, a \texttt{corrid}, a status (30 = EFFECTIVE), and a list of
attributes.

\subsection{PROV-O SDL Full Text}

The \texttt{PROV\_O\_SDL} constant configured by \texttt{configure\_sdl()}:

\begin{lstlisting}
enum DataTypes @datatypeRegistry {
    STRING  @datatype(id: 1)
    TIME    @datatype(id: 2)
    INTEGER @datatype(id: 3)
    LONG    @datatype(id: 4)
    DOUBLE  @datatype(id: 5)
    BOOLEAN @datatype(id: 6)
}

enum Attributes @attributeRegistry {
    prov_generatedAtTime   @attribute(datatype: TIME,   array: false, name: "prov:generatedAtTime",   uri: "http://www.w3.org/ns/prov#generatedAtTime")
    prov_startedAtTime     @attribute(datatype: TIME,   array: false, name: "prov:startedAtTime",     uri: "http://www.w3.org/ns/prov#startedAtTime")
    prov_endedAtTime       @attribute(datatype: TIME,   array: false, name: "prov:endedAtTime",       uri: "http://www.w3.org/ns/prov#endedAtTime")
    prov_invalidatedAtTime @attribute(datatype: TIME,   array: false, name: "prov:invalidatedAtTime", uri: "http://www.w3.org/ns/prov#invalidatedAtTime")
    prov_wasAttributedTo   @attribute(datatype: STRING, array: false, name: "prov:wasAttributedTo",   uri: "http://www.w3.org/ns/prov#wasAttributedTo")
    prov_value             @attribute(datatype: STRING, array: false, name: "prov:value",             uri: "http://www.w3.org/ns/prov#value")
    prov_location          @attribute(datatype: STRING, array: false, name: "prov:location",          uri: "http://www.w3.org/ns/prov#location")
    prov_type              @attribute(datatype: STRING, array: false, name: "prov:type",              uri: "http://www.w3.org/ns/prov#type")

    rdfs_label             @attribute(datatype: STRING, array: false, name: "rdfs:label",             uri: "http://www.w3.org/2000/01/rdf-schema#label")
    rdfs_comment           @attribute(datatype: STRING, array: false, name: "rdfs:comment",           uri: "http://www.w3.org/2000/01/rdf-schema#comment")

    vannak_data_individual @attribute(datatype: STRING, array: false, name: "vannak:dataIndividualId")
    vannak_process_instance @attribute(datatype: STRING, array: false, name: "vannak:processInstanceId")
    vannak_activity_id     @attribute(datatype: STRING, array: false, name: "vannak:activityId")
    vannak_pipeline_id     @attribute(datatype: STRING, array: false, name: "vannak:pipelineId")
    vannak_tenant_id       @attribute(datatype: STRING, array: false, name: "vannak:tenantId")
    vannak_environment_id  @attribute(datatype: STRING, array: false, name: "vannak:environmentId")
}

type ProvEntity @template(name: "ProvEntity") {
    label:             String @use(attribute: rdfs_label)
    comment:           String @use(attribute: rdfs_comment)
    generatedAtTime:   String @use(attribute: prov_generatedAtTime)
    invalidatedAtTime: String @use(attribute: prov_invalidatedAtTime)
    wasAttributedTo:   String @use(attribute: prov_wasAttributedTo)
    value:             String @use(attribute: prov_value)
    location:          String @use(attribute: prov_location)
    type:              String @use(attribute: prov_type)
    dataIndividualId:  String @use(attribute: vannak_data_individual)
}
\end{lstlisting}

\subsection{Integration Test Runbook}

\begin{enumerate}
\item Start PostgreSQL:
  \shellcmd{docker compose -f docker-compose.test.yml up -d}.
\item Run the test:
  \shellcmd{VANNAK_PG_INTEGRATION=1 cargo test --features ipto-writer --test vannak_integration -- --nocapture}.
\item Tear down:
  \shellcmd{docker compose -f docker-compose.test.yml down -v}.
\item Or run everything via
  \shellcmd{./scripts/run-integration-test.sh} (add \shellcmd{--cluster} for the
  \raft{} cluster).
\item Run the Kafka smoke test with
  \shellcmd{./scripts/run-integration-test.sh --kafka-only}, or add
  \shellcmd{--kafka} to the full integration run.
\end{enumerate}

The test exercises the full ingest, index, outbox, and \ipto{} writer flow plus
a shard-range rebalance.

\subsection{Cargo Commands}

\begin{table}[htbp]
\centering
\begin{tabular}{ll}
\hline
Command & Purpose \\
\hline
\shellcmd{cargo build} & build the dependency-free core \\
\shellcmd{cargo test} & run all unit tests \\
\shellcmd{cargo fmt --check} & verify formatting \\
\shellcmd{cargo clippy --all-targets -- -D warnings} & lint with warnings as errors \\
\shellcmd{cargo build --features ipto-writer} & build with the \ipto{} writer \\
\shellcmd{cargo build --features node --bin vannak-node} & build the node binary \\
\hline
\end{tabular}
\caption{Common Cargo commands.}
\end{table}

\subsection{Codec Reference}

\texttt{IptoWritePayload::encode} produces a small, stable, little-endian binary
format used for outbox segment records. Strings are length-prefixed
(\texttt{u32} length followed by UTF-8 bytes).

\begin{lstlisting}[numbers=none]
target:           string
shard_id:         u64
idempotency_key:  string
mapping_version:  string
attribute_count:  u32
repeated attribute_count times:
  attribute_name: string
  value:          tag(u8) + value-body
\end{lstlisting}

Value tags: \texttt{0} String (string), \texttt{1} Integer (\texttt{i64}),
\texttt{2} Boolean (\texttt{u8}), \texttt{3} Timestamp (string), \texttt{4}
StringList (\texttt{u32} count followed by that many strings). Decoding rejects
trailing bytes, unexpected end of input, invalid UTF-8, and unknown value tags.

\subsection{Segment Format}

\begin{lstlisting}[numbers=none]
magic[8]                  // "VANNAK01"
repeated:
  len: u32 little endian
  checksum: u64 little endian
  payload: [u8; len]
\end{lstlisting}

The per-record checksum is an FNV-1a derivative over the payload; the rolling
segment checksum combines records with a rotate-and-xor step.

\subsection{Repository Layout}

\begin{lstlisting}[numbers=none]
src/
  ingest.rs         process event envelope, validation, and journal
  process.rs        process model and state reducer
  data.rs           data-individual provenance model
  metadata.rs       typed metadata references
  durga.rs          Durga monitor compatibility types
  index.rs          dependency-free hot index
  query.rs          query request/result types
  ipto.rs           placement, mapping, payload, outbox model
  ipto_adapter.rs   Ipto writer backed by ipto_rust (feature: ipto-writer)
  storage.rs        append-only segment writer/reader
  cluster.rs        Raft-facing control state and placement
  raft_sm.rs        graft state machine adapter (feature: raft)
  observability.rs  owned snapshot types
  runtime.rs        shard-local runtime, bounded ingest, and Sitas boundary
  sitas_runtime.rs  Sitas executor/mailbox adapter (feature: sitas-runtime)
  bin/vannak-node.rs  Raft node binary (feature: node)
docs/
  ARCHITECTURE.md   detailed architecture notes
\end{lstlisting}

\subsection{Glossary}

\begin{description}
\item[Process event] a \durga{} lifecycle event, ingested as a
  \texttt{PipelineEvent}.
\item[Data individual] a single flowing data item with a stable
  \texttt{DataIndividualId}.
\item[Passive metadata] facts observed at receive/create boundaries.
\item[Active metadata] facts produced by processing (transform, mask, validate,
  enrich).
\item[\texttt{DataIndividualShardId}] domain placement key selecting an \ipto{}
  instance; not a \sitas{} executor shard.
\item[Placement map] an epoch-versioned ring plus overrides, replicated through
  \raft{}.
\item[Outbox] the durable, replayable queue of pending \ipto{} writes.
\item[Segment] an append-only, checksummed durability file with the
  \texttt{VANNAK01} magic.
\item[Checkpoint] a replicated record of consumption progress for recovery.
\item[Writer lease] single-writer authority for an \ipto{} target, fenced by a
  \texttt{LeaseEpoch}.
\item[Hot path] the shard-local ingest and indexing path.
\item[Agreement path] the \raft{}-coordinated control path.
\end{description}

\endinput
